---
title: "Assignment-6"
author: "Ramya Sri Gautham:ramya.srigautham@rutgers.edu"
date: "11/22/2023"
output: html_document
---

<br>

#### Load dataset
```{r message=FALSE}
#loading file and libraries
library(readxl)
library(forecast)
windDT <-read_excel("~/wind-power-production-us.xlsx")
#Time series data 
ts_data <- ts(windDT$Wind_United_States, end=c(2023,2), frequency=4)
#Time series data from 2016 to 2023
windts <- window(ts_data, c(2016))
#Let's recall components present in the data set.
tsdisplay(windts)

```

<br>
* From just the above plots of the Time Series we can see that there is a upward trend in the data over the period of time and also showing signs of seasonality year to year from the regular peaks in the ACF plot.
<br>
<br>

#### Arima Output

```{r}
#perform auto arima on the time series data set
auto_fit <- auto.arima(windts)
auto_fit_Wostepwise <-auto.arima(windts, stepwise=FALSE, approximation=FALSE)
auto_fit
```
<br>
* As per the output, the **ARIMA(0,0,1)(0,1,0)[4] with drift**  model is selected.<br>
* ARIMA(0,0,1): This is the non-seasonal part of the model:<br>
  1. The first number (0) is the order of the AutoRegressive (AR) term. A value of 0 means that the model does not use past values of the series itself to predict future values.<br>
  2. The second number (0) is the order of differencing (I for Integrated). Here, 0 means that the series is assumed to be stationary and does not require differencing to make it stationary.<br>
  3.The third number (1) is the order of the Moving Average (MA) term. This indicates that the model uses one lagged forecast error in predicting future values.<br><br>
* (0,1,0)[4]: This is the seasonal component of the model: <br>  
  1. The first number in the parentheses (0) is the seasonal AR term. It's 0, indicating no seasonal AR component.<br>
  2. The second number (1) is the order of seasonal differencing. This means the data is seasonally differenced once to achieve stationarity.<br>
  3.The third number (0) is the seasonal MA term, which is also 0, indicating no seasonal MA component.<br>
  4.The number in the square brackets ([4]) represents the seasonal period (quarterly data).<br> 
* The coefficients and their standard errors (s.e.) are given for each term. The model's performance is indicated by the log likelihood, AIC, AICc, and BIC values.<br>

```{r}
# Auto arima with stepwise=False
auto_fit_Wostepwise
```

* Here, the **ARIMA(0,0,0)(0,1,1)[4] with drift**  model is selected.<br>
* non seasonal component part of the model contains 0 order of the moving Average.
* seasonal component part contains 1 moving average.
* As the AIC,BIC,accuracy is better in stepwise =False case. we will be considering this model for the prediction.<br>

#### Residual Analysis

```{r}
#Residual Analysis
tsdisplay(residuals(auto_fit_Wostepwise))
```
<br>
* As per Acf plot there are no significant spike. Hence its only white noise.
<br>
```{r}
tsdiag(auto_fit_Wostepwise)
```
<br>
* The Ljung-Box test is applied to the residuals and interpret the results. P-values greater than a chosen significance level indicate that the residuals do not show significant autocorrelation.
<br>
```{r}
hist(auto_fit_Wostepwise$residuals)
```

<br>
* As per histogram the residuals randomly distributed with a mean of zero.
<br>

#### Perform forecast
```{r}
#Forecast for the next 5 period
windts_for <- forecast(auto_fit_Wostepwise, h=5)
plot(windts_for)
```
<br>
* As the drift coefficient of the arima model is **2036.6969** positive ,  the forecast shows the linear trend in upward direction.<br>
* Given the seasonal MA and differencing components, the forecast account for the seasonality observed in the data, typically repeating seasonal patterns observed in the historical data every quarters.<br>
* 95% confidence intervals provides the range within which the future values are expected to lie.
<br>

#### Accuracy of the model
```{r}
# Accuracy of arima with stepwise =TRUE
accuracy(auto_fit)
#Accuracy of arima with stepwise =FALSE
accuracy(auto_fit_Wostepwise)
```

<br>

##### As per the accuracy RMSE of arima model **ARIMA(0,0,0)(0,1,1)[4] with drift** has **6759.059** which is better than the **ARIMA(0,0,1)(0,1,0)[4] with drift**. hence we can forecast using arima model **ARIMA(0,0,0)(0,1,1)[4] with drift**.